---
title: "RMS Titanic"
author: "Jen Wei"
date: 2020-07-13
output:
  github_document:
    toc: true
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- `df_titanic` contains five variables: `Class`, `Sex`, `Age`, `Survived`, `n`

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- According to the [Wikipedia article on the Sinking of the RMS Titanic](https://en.wikipedia.org/wiki/Sinking_of_the_RMS_Titanic#Casualties_and_survivors) (under the "Casualties and Survivors"), the total count of passengers was approximately 1,316, the total count of the crew was 908, and the total count of passengers and crew was 2,224
- Based on the dataset, the total count of passengers was 2,201
- There is trivial difference between the total count of passengers and crew from Wikipedia and that of the dataset (23) (~1%)

Why?
- One guess is that the difference is due to some data scrubbing - from [some Googling](https://www.codementor.io/@innat_2k14/titanic-kaggle-machine-learning-from-disaster-part-1-10gbvqzguc), it seems like this dataset is from Kaggle, and data is often pre-processed to make it usable
- Another guess is that the journey was not direct, so there might've been some miscounts at each stop along the way

Also!
- Interestingly enough, the "Passengers" section of the [original Wikipedia article on the RMS Titanic](https://en.wikipedia.org/wiki/RMS_Titanic#Passengers) mentions "Titanic's passengers numbered approximately 1,317 people" despite supposedly referencing the other one (Sinking of the RMS Titanic)
  - Seems like someone might've misquoted and now makes me a bit weary of trusting the data from these Wikipedia articles

__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task-survived}
df_titanic_survived <- filter(df_titanic, Survived == "Yes")

df_titanic_survived

df_titanic_survived %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge") +
  labs(title = "Survival Count by Class split by Sex")

df_titanic_survived %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge") + facet_wrap(~ Age) +
  labs(title = "Survival Count by Class (split by Sex and Age)")
```

**Observations**:

- There were no children in the crew (which makes sense because that'd be illegal I think otherwise?)
- In `1st`, the majority of the survivors were `Female`
- In `Crew`, the majority of survivors were `Male` though it's likely that they also made up the majority of the crew (can generate more plots to verify this)
  - For some reason, this surprises me (though this is only part of the picture and more plots can be generated to verify this)
- For non-crew passengers, the majority of survivors were women
  - Wonder if there were a lot of "Jack and Rose"-like couples, where the men died to save the women (not really something we can verify from the dataset)
- The split between male and female `child` survivors is close per-class
  - Wondered if the data might show some sort of favoritism, but as-is, it's unclear from the dataset as it doesn't show any data relating to families

``` {r q3-task-deceased}
df_titanic_deceased <- filter(df_titanic, Survived == "No")

df_titanic_deceased %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge") +
  labs(title = "Deceased Count by Class")

df_titanic_deceased %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge") + facet_wrap(~ Age) +
  labs(title = "Deceased Count by Class (split by Age)")
```

**Observations**:

- The children who did not survive were all in `3rd` class
- People in `1st` had the lowest deceased count among the four classes
- `Crew` had the largest decreased count (specifically for `Male` - there were few deceased `Female` for that class)
- Across all classes, `Male` made up the majority of the deceased
- The deceased counts increase from class to class
  - Looking at [a cross-section of the Titanic](https://en.wikipedia.org/wiki/First-class_facilities_of_the_RMS_Titanic#/media/File:Titanic_cutaway_diagram.png), this makes sense as `1st` was closest to the top deck and lifeboats
  - Reading the Wikipedia article (see `Crew` section), it seems like many of the crew were working on the engine or victualling, which were on decks below where most passengers were

``` {r q3-task-extra}
df_titanic %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge") +
  facet_wrap(~ Age) +
  labs(title = "Passenger Count by Class split by Sex")
```
**Observations**:

- `Crew` was the largest `Class` on the Titanic

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(Total = sum(n), Prop = n / Total) %>%
  ungroup()
df_prop
```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-task-q3-visual-replicate}
df_prop_survived <- filter(df_prop, Survived == "Yes")

df_prop_survived

df_prop_survived %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
  facet_grid(Sex ~ Age) +
  labs(title = "Survival Proportions by Class (split by Sex and Age)")
```

**Observations**:

- As noticed in earlier questions of this challenge, the survival of women was disproportionally greater than the survival of men for most adults (i.e. aside from those in `3rd` class)
- Children in `1st` and `2nd` class had high survival rates
- All children from `1st` and `2nd` survived, but less than half the children in `3rd` survived
- Of the surviving adults, aside from `3rd`, more than 50% of the women survived
- Within `Crew`, over 75% of `Female` members survived
  - Wonder which part of the crew they were in - seems plausible that they were part of the Deck department
  - `Male` in `2nd` had the lowest survival rate
- Normalizing the data validates that the proportion of people who survived from `Crew` is not vastly greater than the proportion of other classes
  - When looking at proportions, the survival of `Crew` members is comparable to `2nd` class but is not the highest survival proportion
  - From the count plots in earlier parts of the challenge, it seemed odd that so many more crew members survived than passengers, but it makes sense that the proportion isn't as high
    - Since `Crew` represented the largest `Class` on the Titanic, it seemed like there were a lot of survivors based on the counts, but when looking at proportions, we can see that the survival proportion was not the greatest among the classes

__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!
  
```{r q5-task}
df_survived <- filter(df_titanic, Survived == "Yes")

total_survived <- sum(df_survived$n)

df_survived_group_prop <-
  df_survived %>%
  group_by(Class, Sex, Age) %>%
  mutate(Group_Prop = n / total_survived) %>%
  ungroup()

df_survived_group_prop

df_survived_group_prop %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Group_Prop, fill = Sex)) +
  facet_grid(Sex ~ Age) +
  labs(title = "Survival Group Proportions by Class (split by Sex and Age)")
```

**Observations**:

- Among all the survivors, `Male` survivors from `Crew` made up the largest percentage (> 25%)
- Among all the passengers, `Female` survivors from `1st` made up the largest percentage (~20%)
- Children made up a small percentage of the survivors
  - This is not unexpected given how few children there were aboard the Titanic

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
