---
title: "Gapminder"
author: "Jen Wei"
date: 2020-07-24
output:
  github_document:
    toc: true
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to education people about the state of the world. They promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

__q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
glimpse(gapminder)
summary(gapminder)
```

**Observations**:

- The variables in the dataset include `country`, `continent`, `year`, `lifeExp`, `pop`, and `gdpPercap`

__q1__ Determine the most and least recent years in the `gapminder` dataset.

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(select(gapminder, year))
year_min <- min(select(gapminder, year))

year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)
print("Nice!")
```

__q2__ Filter on years matching `year_min`, and make a plot of the GDE per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
year_min_df <- filter(gapminder, year == year_min)
glimpse(year_min_df)

year_min_df %>%
  ggplot(mapping = aes(y = gdpPercap, x = continent)) +
  geom_boxplot() +
  labs(title = "GDP per capita by continent in 1952") +
  scale_y_log10()
```

**Observations**:

- Asia has the largest range of GDP per capita
- Oceania has the smallest range of GDP per capita
- Asia has the country with the largest GDP per capita (though it's an outlier)
  - Excluding the outliers, Europe has the country with the largest GDP per capita
- Africa has the country with the lowest GDP per capita
- Oceania has the largest median GDP per capita
- Africa has the smallest median GDP per capita
- There are three outliers identified in the dataset - two in Americas and one in Asia

**Difficulties & Approaches**:

- For visualizing `gdpPercap` vs `continent`, I first tried a column plot, but it was a bit noisy (as there seemed to be a line per row in the filtered dataset), and the main takeaway from it was the total count per continent. Then, I decided to use a boxplot to see trends within each `continent`. This worked better, but the spread of data across countries and continents made certain boxplots squished. To stretch them out a bit, I log-scaled `gdpPercap`.

__q3__ You should have found at least three outliers in q2. Identify those outliers (figure out which countries they are).

```{r q3-task}
americas_outliers_df <- filter(year_min_df, gdpPercap > 1.000e+04, continent == 'Americas')
glimpse(americas_outliers_df)

asia_outliers_df <- filter(year_min_df, gdpPercap > 1.000e+05, continent == 'Asia')
glimpse(asia_outliers_df)
```

**Observations**:

I initially thought of using the plot to identify the outliers via [this StackOverflow post](https://stackoverflow.com/questions/33524669/labeling-outliers-of-boxplots-in-r), but then I realized I could just filter the data based on info from the q2 plot of GDP per capita by continent in 1952.

- The outliers are Canada, United States, and Kuwait

NOTE: This was done manually and could've been done in a smarter way, but it works?

__q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
year_extremes_df <- filter(gapminder, year == year_min | year == year_max)
glimpse(year_extremes_df)

year_extremes_df %>%
  ggplot(mapping = aes(y = gdpPercap, x = continent, color = factor(year))) +
  geom_boxplot() +
  facet_grid(~ year) +
  stat_summary(
    fun = median,
    geom = "line",
    aes(group = year, color = factor(year)),
    lwd = 1,
    position = position_dodge(width = .75)
  ) +
  scale_y_log10() +
  labs(title = "GDP per capita by continent")
```

**Observations**:
- GDP per capita by continent (from highest to lowest) is the same for the earliest and latest years
  - Oceania, Europe, Americas, Asia, Africa
- The spread of GDP per capita across countries for each continent has increased between 1952 and 2007
- Both years have three outliers
  - 2007 outliers are all in Americas: Haiti, Canada, and United States
  
``` {r q4-task-outliers}
americas_latest_outliers_df <- filter(year_extremes_df, gdpPercap > 2.000e+04 | gdpPercap < 2.500e+03 , continent == 'Americas', year == 2007)
glimpse(americas_latest_outliers_df)
```

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

__q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

While going through EDA for GDP per capita, I was curious about how other variables impact GDP per capita, so first, I'll look into population.

```{r q5-task1}
gapminder %>%
  ggplot(mapping = aes(x = pop, y = gdpPercap, color = year)) +
  geom_point() +
  facet_wrap(~ continent, nrow = 5) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "GDP per capita by population per continent over time")
```

**Observations:**
- Seems like there are some groupings within the data that show a linear, positive trend
  - Over time, both population and GDP per capita increased for Oceania
    - There are two distinct groups, one for each Oceanian country (New Zealand and Australia)

There's a lot of data to digest here, so similar to the guided EDA, I'll now focus on the earliest and latest years, 1952 and 2007 and try to familiarize myself with just population.

```{r q5-task2}
year_extremes_df %>%
  ggplot(mapping = aes(y = pop, x = continent, color = factor(year))) +
  geom_boxplot() +
  facet_grid(~ year) +
  stat_summary(
    fun = median,
    geom = "line",
    aes(group = year, color = factor(year)),
    lwd = 1,
    position = position_dodge(width = .75)
  ) +
  scale_y_log10() +
  labs(title = "Population by continent for the year-extremes")

```

**Observations:**

- The median population size increased for all continents between 1952 and 2007
- The largest median population size is associated with Asia in both years
- There is only one outlier in 2007 compared to three in 1952
- The median population sizes for Africa, Americas, Europe, and Oceania look similar on the log scale (or at least less distinct) for 2007 compared to 1952

The insights from the plot are expected as the world population is consistently growing, and this served as a helpful check to ensure there aren't any abnormalities before looking at population AND GDP per capita.

```{r q5-task3}
# Needed for geom_label_repel
library(ggrepel)

year_extremes_min_max_df <-
  year_extremes_df %>%
  group_by(continent, year) %>%
  filter(pop == min(pop) | pop == max(pop))

year_extremes_min_max_df

year_extremes_df %>%
  ggplot(mapping = aes(x = pop, y = gdpPercap, color = continent)) +
  geom_point() +
  geom_label_repel(
    data = year_extremes_min_max_df,
    aes(label = gdpPercap)
  ) +
  facet_grid(~ year ~ .) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "GDP per capita by population per continent for 1952 and 2007 (with first and last population counts annotated)")
```

**Observations:**

Looking at the extremes of each continent in 1952, the smallest populated countries in Africa and the Americas had lower GDP per capita than that of the largest populated countries, while in Asia, Europe, and Oceania, the opposite is true where the smallest populated countries had higher GDP per capita than that of the largest populated countries.

  - In Africa, the smallest populated country had a *lower* GDP per capita than that of the largest populated country 
  - In Americas, the smallest populated country had a *lower* GDP per capita than that of the largest populated country
  - In Asia, the smallest populated country had a *higher* GDP per capita than that of the largest populated country
  - In Europe, the smallest populated country had a *higher* GDP per capita than that of the largest populated country
  - In Oceania, the smallest populated country had a *higher* GDP per capita than that of the largest populated country

Looking at the extremes of each continent in 2007, all countries had the same relations between the extremes aside from Oceania, where the smallest populated country had a *lower* GDP per capita than that of the largest populated country (which is different from 1952).

Both generally and by continent, there is no linear relationship between population and GDP per capita. This is not totally unexpected though as each point is a country, and each country has its own economy, and there are other factors not represented in the dataset that likely more directly and more greatly influenced GDP per capita.

**Bonus!**

After chatting with the rest of Team Zeta, people were curious about drilling into the second plot of GDP per capita and population per continent. Thought about doing that earlier but ran out of time/energy. However, I ended up circling back and figured I'd add it here.

```{r q5-task4}
gap_minder_mean_df <- gapminder %>%
  group_by(continent, year) %>%
  summarise(mean_gdpPercap = mean(gdpPercap), mean_pop = mean(pop))

gap_minder_mean_df

gap_minder_mean_df %>%
  ggplot(mapping = aes(x = mean_pop, y = mean_gdpPercap, color = year)) +
  geom_point() +
  facet_wrap(~ continent, nrow = 5) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Mean GDP per capita by mean population per continent over time")
```
**Observations:**

When looking at the mean population and mean GDP per capita on a log-log scale, we can see a somewhat linear trend of population and GDP per capita both increasing over the years.

- Europe has the smallest mean population range over time
- Africa has the largest mean population range over time
- Slope of the Europe datapoints is the sharpest across the five continents
