---
title: "Gapminder"
author: "Jen Wei"
date: 2020-07-24
output:
  github_document:
    toc: true
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to education people about the state of the world. They promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

__q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
glimpse(gapminder)
summary(gapminder)
```

**Observations**:

- The variables in the dataset include `country`, `continent`, `year`, `lifeExp`, `pop`, and `gdpPercap`

__q1__ Determine the most and least recent years in the `gapminder` dataset.

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(select(gapminder, year))
year_min <- min(select(gapminder, year))

year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)
print("Nice!")
```

__q2__ Filter on years matching `year_min`, and make a plot of the GDE per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
year_min_df <- filter(gapminder, year == year_min)
glimpse(year_min_df)

year_min_df %>%
  ggplot(mapping = aes(y = gdpPercap, x = continent)) +
  geom_boxplot() +
  labs(title = "GDP per capita by continent in 1952") +
  coord_flip() +
  scale_y_log10()
```

**Observations**:

- Asia has the largest range of GDP per capita
- Oceania has the smallest range of GDP per capita
- Asia has the country with the largest GDP per capita (though it's an outlier)
  - Excluding the outliers, Europe has the country with the largest GDP per capita
- Africa has the country with the lowest GDP per capita
- Oceania has the largest median GDP per capita
- Africa has the smallest median GDP per capita
- There are three outliers identified in the dataset - two in Americas and one in Asia

**Difficulties & Approaches**:

- For visualizing `gdpPercap` vs `continent`, I first tried a column plot, but it was a bit noisy (as there seemed to be a line per row in the filtered dataset), and the main takeaway from it was the total count per continent. Then, I decided to use a boxplot to see trends within each `continent`. This worked better, but the spread of data across countries and continents made certain boxplots squished. To stretch them out a bit, I applied `coord_flip()`, but it was still squished, so I log-scaled `gdpPercap`.

__q3__ You should have found at least three outliers in q2. Identify those outliers (figure out which countries they are).

```{r q3-task}
americas_outliers_df <- filter(year_min_df, gdpPercap > 1.000e+04, continent == 'Americas')
glimpse(americas_outliers_df)

asia_outliers_df <- filter(year_min_df, gdpPercap > 1.000e+05, continent == 'Asia')
glimpse(asia_outliers_df)

```

**Observations**:

I initially thought of using the plot to identify the outliers via [this StackOverflow post](https://stackoverflow.com/questions/33524669/labeling-outliers-of-boxplots-in-r), but then I realized I could just filter the data based on data gleaned from the q2 plot of GDP per capita by continent in 1952.

- The outliers are Canada, United States, and Kuwait

NOTE: I probably could've just swapped x and y instead of using `coord_flip()`, but alas

__q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
year_extremes_df <- filter(gapminder, year == year_min | year == year_max)
glimpse(year_extremes_df)

year_extremes_df %>%
  ggplot(mapping = aes(x = gdpPercap, y = continent, color = factor(year))) +
  geom_boxplot() +
  facet_grid(year ~ .) +
  stat_summary(
    fun = median,
    geom = "line",
    aes(group = year, color = factor(year)),
    lwd = 1,
    position = position_dodge(width = .75)
  ) +
  scale_x_log10() +
  labs(title = "GDP per capita by continent")

```

**Observations**:
- GDP per capita by continent (from highest to lowest) is the same for the earliest and latest years
  - Oceania, Europe, Americas, Asia, Africa
- The spread of GDP per capita across countries for each continent has increased between 1952 and 2007
- Both years have three outliers
  - 2007 outliers are all in Americas: Haiti, Canada, and United States
  
``` {r q4-task-outliers}
americas_latest_outliers_df <- filter(year_extremes_df, gdpPercap > 2.000e+04 | gdpPercap < 2.500e+03 , continent == 'Americas', year == 2007)
glimpse(americas_latest_outliers_df)
```

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

__q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
```


```{r q5-task2}
## TASK: Your second graph
```


```{r q5-task3}
## TASK: Your third graph
```

