---
title: "Food Consumption and CO2 Emissions"
author: "Jen Wei"
date: '2020-08-03'
output:
  github_document:
    toc: yes
  html_notebook:
    toc: yes
---

*Purpose*: <TODO>

```{r setup}
library(tidyverse)
```

*Background*: <TODO>

# Get the Data
<!-- -------------------------------------------------- -->

```{r final-project-get-data}

df_food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')

glimpse(df_food_consumption)

summary(df_food_consumption)
```

# Analyze
<!-- -------------------------------------------------- -->

```{r EDA-analyze-plot-1}
df_food_consumption %>%
  ggplot(mapping = aes(x = consumption, y = co2_emmission)) +
  geom_point(mapping = aes(color = food_category)) +
  # scale_y_log10(labels = scales::label_number_si()) +
  labs(y = "CO2 Emission", x = "Consumption", title = "CO2 Emission vs Consumption (per capita) by Food Category")
```

**Observations:**

- Linear relationships between CO_2 emission and consumption across all food categories
- "Beef" has an outlier for CO_2 emission
- "Milk and cheese" has an outlier for consumption

From the observations, I wondered who the highest consumers were for each food category

```{r EDA-analyze-plot-2}
df_food_consumption %>%
  ggplot(mapping = aes(x = food_category, y = consumption)) +
  geom_boxplot(mapping = aes(color = food_category)) +
  labs(x = "Food Category", y = "Consumption (kg/person/year)", title = "Consumption (per capita) by Food Category") +
  geom_text( 
    data = df_food_consumption %>% group_by(food_category) %>% filter(consumption == max(consumption)),
    mapping = aes(label = country, fill = "black"),
    size = 3,
    nudge_x = 0.5,
    vjust = "inward",
    hjust = "inward",
    check_overlap = TRUE
  ) +
  coord_flip()
```

**Observations:**

Per capita . . .
- Tunisia is the largest cosumer of wheat
- Taiwan is an outlier and the largest consumer of soybeans
- Bangladesh is the largest consumer of rice
- Israel is the largest consumer of poultry
- Hong Kong SAR China is the largest consumer of pork
- UAE is the largest consumer of nuts and peanut butter
- Finland is an outlier and the largest consumer of milk and cheese
- Iceland is the largest consumer of lamb and goat
- Maldives is an outlier and the largest consumer of fish
  - Despite their size, the Maldives is "99% sea" (according to goway.com), where fish is a primary part of their diet
- Japan is the largest consumer of eggs
- Argentina is an outlier and the largest consumer of beef

```{r EDA-analyze-plot-3}
df_food_consumption %>%
  ggplot(mapping = aes(x = food_category, y = co2_emmission)) +
  geom_boxplot(mapping = aes(color = food_category)) +
  labs(x = "Food Category", y = "CO2 Emission (kg/person/year)", title = "CO2 Emission by Food Category") +
  geom_text( 
    data = df_food_consumption %>% group_by(food_category) %>% filter(co2_emmission == max(co2_emmission)),
    mapping = aes(label = country, fill = "black"),
    size = 3,
    nudge_x = 0.5,
    vjust = "inward",
    hjust = "inward",
    check_overlap = TRUE
  ) +
  coord_flip()
```

**Observations:**

- Similar to the last plot (since there's a positive relationship between consumption and CO_2 emission)
- Most meats (Beef, Lamb & Goat, Pork, Fish) have CO_2 emission values around or above 250 kg/person/year (aside from poultry)
- Rice is the largest CO_2 emitter aside from meats

## Team Exploration

From initial exploration, we decided to focus on BRIC countries and how the United States compares

```{r team-bricu-plot-1}
df_food_consumption_bricu <- df_food_consumption %>%
  filter(country == 'Brazil' | country == 'China' | country == 'India' | country == 'Russia' | country == 'United States')

glimpse(df_food_consumption_bricu)

df_food_consumption_bricu %>%
  ggplot(mapping = aes(x = food_category, y = co2_emmission)) +
  geom_point(mapping = aes(color = country)) +
  scale_y_log10(labels = scales::label_number_si()) +
  labs(x = "Food Category", y = "CO2 Emission (kg/person/year)", title = "CO2 Emission by Food Category") +
  coord_flip()
```

**Observations:**

- TODO

## Extra Unused "Code"

```{r exploration-data-wrangling}

# Find country with max emission for each food category
df_max_emission <- df_food_consumption %>%
  group_by(food_category) %>%
  summarize(
    emission = max(co2_emmission),
    country = country[which.max(co2_emmission)],
    consumption = consumption[which.max(co2_emmission)],
    food_category = food_category
  ) %>%
  distinct()

# Find country with min emission for each food category
df_min_emission <- df_food_consumption %>%
  group_by(food_category) %>%
  summarize(
    emission = min(co2_emmission),
    country = country[which.min(co2_emmission)],
    consumption = consumption[which.min(co2_emmission)],
    food_category = food_category
  ) %>%
  distinct()

# Find country with max food consumption for each food category
df_max_consumption <- df_food_consumption %>%
  group_by(food_category) %>%
  summarize(
    emission = max(consumption),
    country = country[which.max(consumption)],
    consumption = consumption[which.max(consumption)],
    food_category = food_category
  ) %>%
  distinct()

# Find country with min consumption for each food category
df_min_consumption <- df_food_consumption %>%
  group_by(food_category) %>%
  summarize(
    emission = min(consumption),
    country = country[which.min(consumption)],
    consumption = consumption[which.min(consumption)],
    food_category = food_category
  ) %>%
  distinct()

# Merge all the max/min dataframes into one
df_emission_consumption_extremes <- merge(df_max_emission,df_min_emission, all = TRUE) %>%
  merge(df_max_consumption, all = TRUE) %>%
  merge(df_min_consumption, all = TRUE)
```
```{r exploration-plot}
df_emission_consumption_extremes %>%
  ggplot(mapping = aes(x = food_category, y = emission)) +
  geom_point(mapping = aes(color = food_category)) +
  scale_y_log10(labels = scales::label_number_si()) +
  labs(y = "CO2 Emission", x = "Consumption", title = "Max/Min CO2 Emission vs Consumption by Food Category") +
  geom_text(aes(label = country), check_overlap = TRUE)
```

# Notes / Sources
<!-- -------------------------------------------------- -->

[1] <TODO>
